Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-16 paper/agree.tex
Revise the AGREE semantics and reorder section

This section introduces the AGREE contract specification language through the cyber-hardened system of \figref{fig:hardened}.
It details the specification of that top-level system interface.
It then discusses the specifications for the filter and the monitor added by the transforms.

This section then formalizes AGREE's assume-guarantee reasoning by defining the set of verification conditions that AGREE must check.
The verification conditions succinctly summarize what in means when AGREE proves that a system implementation satisfies it specification.
These conditions are useful in understanding how the semantics in synthesis differ from those in the assume-guarantee reasoning in \secref{sec:semantics}.

\subsection{Contract Specification}
\newsavebox{\sw}
\begin{lrbox}{\sw}
\begin{lstlisting}[style=agree,numbers=left]
eq req : bool = event(AutomationRequest);
eq avl : bool = event(AirVehicleLocation);
eq wp : bool = event(Waypoint);
eq strt: bool = event(Start);
eq alrt : bool = event(Alert);

assume "Automation requests are well-formed" :
  req => WELL_FORMED_AUTOMATION_REQUEST(AutomationRequest);
assume "Air vehicle locations are well-formed" :
  avl => WELL_FORMED_WAYPOINT(AirVehicleLocation);    
assume "One automation request in flight at a time" :
  true -> 
  (req => pre(Historically(not req) or Since(not req, strt)));

@@

The AGREE specification for the SW component in the example of Section~\ref{sec:example} with the added cyber requirements is given in \figref{fig:sw}.
The AGREE specification language is a first-order predicate calculus that uses stream concepts, and operators, from the Lustre language \cite{10.1145/41625.41641}.
As with Lustre, the semantics are synchronous dataflow where the inputs, outputs, and expressions are data streams that comply with the input assumptions.
Contracts are evaluated in dependency order with inputs being propagated to outputs through all contracts until they stabilize; as such, the contracts, and thereby the top-level model, must be acyclic.\footnote{An apparent syntactic cycle, where a component is linked back to itself, may be broken temporally by inserting delay elements.}
Once the contracts have stabilized, the model takes a synchronous step to the next input data in the stream.
The semantics do not model computation or communication delay.
The output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

The AGREE model checker attempts to prove several properties of the top-level model being verified.
The first is that the output guarantees of each component implementing the system are strong enough to validate the input assumptions of any downstream component as well as to satisfy the guarantees of the output of the top-level component being verified (\ie, the system composition meets input assumptions at each input as well as the guarantees on the system output).
These properties are reported in the expanded lists in \figref{fig:example}(b) and \figref{fig:hardened}(b).
The next set of properties prove that the contract specifications for each component are \emph{self-consistent} \ie, a contract does not contradict itself.
These are the unexpanded results at the bottom of the figures.

Returning back to the contract of \figref{fig:sw}, it uses \texttt{eq} statements to define variables local to the contract specification.
For example, the \texttt{req} variable is equivalent to the \texttt{event(AutomationRequest)} expression.
In the AGREE semantics, there is an implicit \emph{event} input (or output) associated with every named event port in a component.
The semantics used here do not buffer these events so the implicit input (or output) is only a boolean value.
An \texttt{event} expression refers to that implicit input (or output) and is true when data is placed on the named port.
The system contract here states assumptions on well-formed input, followed by guarantees on properties about the output.

>>>>

This section formalizes AGREE's assume-guarantee reasoning by defining the set of verification conditions that AGREE must check.
The verification conditions succinctly summarize what in means when AGREE proves that a system implementation satisfies it specification.
These conditions are useful in understanding how the semantics in synthesis differ from those in the assume-guarantee reasoning in \secref{sec:semantics}.

This section then introduces the AGREE contract specification language through the cyber-hardened system of \figref{fig:hardened}.
It details the specification of that top-level system interface.
It then discusses the specifications for the filter and the monitor.
These two specifications are those that are synthesized to equivalent CakeML by SPLAT.

\subsection{Contract Language Specification}
\newsavebox{\sw}
\begin{lrbox}{\sw}
\begin{lstlisting}[style=agree,numbers=left,escapeinside={*}{*}]
eq req : bool = event(AutomationRequest);*\label{line:sw-event-def-start}*
eq avl : bool = event(AirVehicleLocation);
eq wp : bool = event(Waypoint);
eq strt: bool = event(Start);
eq alrt : bool = event(Alert);*\label{line:sw-event-def-end}*

assume "Automation requests are well-formed" : *\label{line:sw-assume-1}*
  req => WELL_FORMED_AUTOMATION_REQUEST(AutomationRequest);
assume "Air vehicle locations are well-formed" : *\label{line:sw-assume-2}*
  avl => WELL_FORMED_WAYPOINT(AirVehicleLocation);    
assume "One automation request in flight at a time" : *\label{line:sw-assume-3}*
  true -> 
  (req => pre(Historically(not req) or Since(not req, strt)));

@@

The AGREE specification language uses stream concepts, and operators, from the Lustre language \cite{10.1145/41625.41641}.
As with Lustre, the semantics are synchronous dataflow where the inputs, outputs, and expressions are data streams that comply with the input assumptions.
Contracts are evaluated in dependency order with inputs being propagated to outputs through all contracts until they stabilize; as such, the contracts, and thereby the top-level model, must be acyclic.\footnote{An apparent syntactic cycle, where a component is linked back to itself, may be broken temporally by inserting delay elements.}
Once the contracts have stabilized, the model takes a synchronous step to the next input data in the stream.
The semantics do not model computation or communication delay.
The output of one contract is seen at the input of any downstream contract in the same step of the input data stream.
The language is best introduced through example.

The AGREE specification for the SW component in the example of Section~\ref{sec:example} is given in \figref{fig:sw}.
The specification uses \texttt{eq} statements to define variables local to the contract specification.
For example, \lineref{line:sw-event-def-start} defines the \texttt{req} variable to be equivalent to the \texttt{event(AutomationRequest)} expression.

All the named ports in the corresponding AADL component are in the scope of the specification, and there are additional implicit boolean \emph{event} inputs (or outputs) associated with event ports.
An \texttt{event} expression refers to that implicit input (or output) boolean value and is true when data is placed on the named port and false otherwise.
\linesref{line:sw-event-def-start}{line:sw-event-def-end} create local variables that are true when data is present on the corresponding event ports for the component.
The local variables here are purely for convenience in writing the specification.

Assumptions are best understood as being evaluated on the pre-state of a component while guarantees are evaluated on the post-state.
Contract languages such as Eifle and Dafny make this pre-state assume evaluation and post-state guarantee evaluation part of the semantics, and it relies on operators such as \texttt{old} it indicate when a guarantee should use a pre-state value in evaluating a guarantee.
AGREE differs from these languages in that it evaluates both the assumptions and guarantees in the post-state of the component.
But like other languages it relies on the operator \texttt{pre} to refer to a pre-state evaluation in both the assumptions and guarantees.
Since streams are associated with all expressions in AGREE meaning that they are defined through time, the \texttt{pre} operator simply returns the previous value of the enclosed expression (e.g., it looks back one step in time).

The \texttt{assume} statement is a string description followed by a predicate.
Those on \lineref{line:sw-assume-1} and \lineref{line:sw-assume-2} are implications requiring that when data is present it is well-formed.
The well-formed predicates themselves are defined elsewhere using AGREE functions.
The assumption on \lineref{line:sw-assume-3} uses the \emph{followed-by} operator, \texttt{->}, and the \texttt{pre} operator to constrain when requests can arrive at the input.

The followed-by expression defines what happens at the first instance in the stream, time 0, and then what follows after. 
Here the assumption is \texttt{true} at time 0, and then takes on the truth value of the implication in all future instances.
The implication uses the \texttt{pre} expression in its left operand. 
In the example, it is guarded by the followed-by since expressions are undefined before time 0.
So in the first instance at time 0, the assumption is \texttt{true}, and then at every instance after, it depends on when the request arrives.

The stream semantics in AGREE mean that it is possible to use PLTL operators that look back in time.
